Antisense transcripts can quickly evolve to encode proteins

### 2023-03-09 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/Divergence_dmel.svg 	modified:   Manuscripts/Figures/M2_main/Divergence_scer.svg 	new file:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerGC.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Manuscripts/M2/refs.bib 	modified:   Modeling/antisenseGenes.jl

We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs. For any stretch of DNA to be an ORF, its sequence should contain 3$n$ nucleotides (with $n$ being at least 3), with a start codon that marks its beginning, and exactly one stop codon that marks its end. Specifically, the absence of any stop codons within the DNA sequence is the most important factor in determining the likelihood of an ORF to exist. That is because the likelihood of a premature stop codon increases exponentially with the ORF's length, whereas likelihood of a start codon and a terminal stop codon are unaffected by the ORF's length.

We estimate the gain and loss probabilities of antisense ORFs using the existing (sense) ORF as a reference. Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost. Antisense ORFs can overlap with the sense ORFs in three different frames. In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF. In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively. In frame 0, the sequence of an antisense codon is determined by the sequence of the exactly overlapping sense codon. However, for frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons.

@@

We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data. Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}. We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}. \autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.

>>>>

We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs. These models are defined by two kinds of probability. The first is the probability of finding a certain kind of DNA sequence, for example an ORF. This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC content or by the frequencies of short DNA sequences (k-mers). The second kind of probability describes the mutational change of a sequence to a different kind of sequence. For example, gain or loss of an ORF. This transition probability depends on the mutation rate and mutation bias, in addition to nucleotide composition. We estimate these parameters primarily from the data on the yeast, \textit{Saccharomyces cerevisiae} \citep{scermutrate}. Our choice is motivated by the fact that yeast is a convenient model organism for laboratory experimental studies that can be used to validate theoretical predictions. We also perform analogous analyses using data obtained from \textit{Drosophila melanogaster} \citep[][Supplementary section XX]{drosophilamutrate}.

We estimate the stationary and transition probabilities of antisense ORFs using the existing (sense) ORF as a reference. Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost. Antisense ORFs can overlap with the sense ORFs in three different frames. In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF. In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively. Thus in frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons (dicodons). Due to this sequence overlap, the evolution of antisense ORFs would be constrained by the evolutionary selection pressures on the sense ORF. Furthermore, these constraints would be different for antisense ORFs located in the three different frames. We analysed the evolution of antisense ORFs when the sense ORF is under three different levels of purifying selection. The first level describes an absence of purifying selection, where any kind of mutation except a non-sense mutation (gain of stop codon) in the sense ORF is tolerated. The second level describes a weak purifying selection that allows synonymous mutations (nucleotide mutations that do not change the encoded amino acid), as well as mutations where an amino acid is substituted by a chemically similar amino acid (for example, aspartic acid to glutamic acid). Finally, in the third level that describes a strong purifying selection, only synonymous mutations are tolerated in the sense ORF.

\subsection*{Antisense ORFs are more likely exist in frame 1}

For any stretch of DNA to be an ORF, its sequence should contain 3$n$ nucleotides (with $n$ being at least 3), with a start codon that marks its beginning, and exactly one stop codon that marks its end. The absence of any stop codon within the DNA sequence is the most important factor in determining the existence of an ORF. That is because the likelihood of a premature stop codon increases exponentially with the ORF's length, whereas the likelihoods of a start codon and a terminal stop codon are independent of the ORF's length. Furthermore, overlapping DNA region between the sense and antisense ORF would mostly consist of non-terminal (internal) codons.

Based on these considerations, we determined the probability of finding an antisense ORF. To this end, we calculated the probability of finding a stop codon the three antisense frames, given the nucleotide composition of sense ORF including the condition that no stop codon exists in the sense ORF. Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF. It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons, respectively. We thus calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons. We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF. Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames. We did so for four different values of GC content. We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases. The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%. Even for these exceptional cases, the probability of an antisense ORF in frame 1 was no less than 91\% of the corresponding ORF probabilities in other frames.

To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths. We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition. We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 were more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs had less than 39 codons and a GC content of 60\%.

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome.

@@
